### 2022-10-06 manuscript/manuscript.tex
more theory

The change in a neutral allele's frequency in one generation can be partitioned
into the three sources of stochasticity: the random associations with fitness
backgrounds (i.e. \emph{draft}), the non-heritable randomness in family size,
and the Mendelian noise from heterozygotes segregating. If we let $x_{0,i} \in
\{0, \nicefrac{1}{2}, 1\}$ be the frequency of neutral alleles individual $i$
in generation 0 carries, we can partition the neutral allele frequency change
into the underlying stochastic causes,

\begin{align}
  p_1 = \frac{1}{2N} \sum_{i=1}^N \left( x_{0,i}k_{0,i} + \sum_{j=1}^{k_{0,i}} \delta_{0,i,j} \right)
\end{align}

@@

Then, noting that $p_0 = \nicefrac{1}{N} \sum_{i=1}^N x_{0,i}$, the allele
frequency change is,

\begin{align}
  \Delta p_1 = p_1 - p_0 &= \frac{1}{2N} \sum_{i=1}^N x_{0,i} (k_{0,i} - 1) + \frac{1}{2N} \sum_{i=1}^N \sum_{j=1}^{k_{0,i}} \delta_{0,i,j} \\
  \Delta p_1 &= K_1 + H_1
\end{align}

@@

Now, let us look the variance of $\var(\Delta p_1)$ over evolutionary
replicates. Since the Mendelian random process and the offspring random process
are independent,

\begin{align}
  \var(\Delta p_1) &= \var(K_1) + \var(H_1).
\end{align}

@@

>>>>

The change in a neutral allele's frequency in one generation can be partitioned
into the three sources of stochasticity: the random associations with fitness
backgrounds (i.e. \emph{draft}), the non-heritable randomness in family size,
and the Mendelian noise from heterozygotes segregating. If we let $x_{0,i} \in
\{0, \nicefrac{1}{2}, 1\}$ be the frequency of neutral alleles individual $i$
in generation 0 carries, we can partition the random neutral allele frequency
of the population ($x_t$ without the individual index) into the underlying
stochastic causes,

\begin{align}
  x_1 = \frac{1}{2N} \sum_{i=1}^N \left( x_{0,i}k_{0,i} + \sum_{j=1}^{k_{0,i}} \delta_{0,i,j} \right)
\end{align}

@@

The frequency in the initial generation is $x_0 = \nicefrac{1}{N} \sum_{i=1}^N
x_{0,i}$, though to indicate that we treat this as fixed, we use $p_0 = x_0$.
Then, the allele frequency change is,

\begin{align}
  \Delta x_1 = x_1 - p_0 &= \frac{1}{2N} \sum_{i=1}^N x_{0,i} (k_{0,i} - 1) + \frac{1}{2N} \sum_{i=1}^N \sum_{j=1}^{k_{0,i}} \delta_{0,i,j} \\
  \Delta x_1 &= K_1 + H_1
\end{align}

@@

Now, let us look the variance of $\var(\Delta x_1)$ over evolutionary
replicates. Since the Mendelian random process and the offspring random process
are independent,

\begin{align}
  \var(\Delta x_1) &= \var(K_1) + \var(H_1).
\end{align}

@@

Now, we look at this variance equation behaves with $t > 1$.

\begin{align}
  \V(x_2) &= \E\left[(x_2 - x_0)^2\right] \\
          &= \E_1\left[\E_2\left[((x_2-x_1) + (x_1 - x_0))^2 | x_1\right]\right] \\
          &= \E_1\left[\E_2\left[(\Delta x_2 + \Delta x_1)^2 | x_1\right]\right] \\
          &= \E_1\left[\E_2\left[{\Delta x_2}^2|x_1\right]\right] + 2\E_1\left[\E_2[\Delta x_2 \Delta x_1 | x_1]\right] + \E_1\left[ {\Delta x_1}^2\right] \\
          &= \E_1\left[\E_2\left[{\Delta x_2}^2|x_1\right]\right] + \E_1\left[ {\Delta x_1}^2\right] + C_{1,2} \\
          &= \frac{p_0(1-p_0)}{2N}\left(1 - \frac{1}{2N}\right) + \frac{p_0(1-p_0)}{2N} + C_{1,2}
\end{align}

Consequently, the variance of the neutral allele frequency changes each
generation according to the probability of failing to coalesce each generation
and the pairwise covariance terms $C_{i,j}$ that build up due to associations
with heritable fitness backgrounds. Note that the covariance term $C_{1,2} =
\E_1\left[\E_2[\Delta x_2 \Delta x_1 | x_1]\right] = 0$ under neutral
evolution, and this simplifies to the well-known equation for variance in a
Wright--Fisher population.

% In general, the variance in allele frequency after $t$ generations
% of neutral evolution is,

%\begin{align}
%  \label{eq:var_freq}
%  %\V(x_t) &= \frac{x_0(1-x_0)}{2N}\left(\sum_{i=1}^t \left(1 - \frac{1}{2N}\right)^{i-1} \right)
%  \V(x_t) &= \V(x_1) \left(1 + \sum_{i=1}^t \left(1 - \frac{1}{2N}\right)^{i} \right)
%\end{align}
